Component response rate variation drives stability in large complex systems

### 2018-05-18 ms.Rmd
New paragraphs

```{r echo = FALSE}
S     <- 200;
C     <- 0.05
sigma <- 0.4;
pr_st <- read.csv(file = "sim_results/bi_gamma/bi_pr_st.csv");
pr_st <- pr_st[,-1];
```

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supporting Information for other distributions).

<!--- Say that the amnat paper did this with multiplying a row --->

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[fhalf]), digits = 3)` centred at $\gamma = -1.95$ and one small circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[shalf]), digits = 3)` centred at $\gamma = -0.05$. In **a**, the maximum real eigenvalaue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in **b** $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in **b** but not **a** is stable because in **b** $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` were stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 1$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = d = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but clusters eigenvalues toward the distribution's centre (-1, 0). Black and red elipses in both panels show the circle centred on the distribution in panels a and b, respectively, which have a radius of $\sigma \sqrt{SC}$. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.

>>>>

```{r echo = FALSE}
S     <- 200;
C     <- 0.05;
sigma <- 0.4;
pr_st <- read.csv(file = "sim_results/bi_gamma/bi_pr_st.csv");
pr_st <- pr_st[,-1];
```

@@

```{r echo = FALSE}
S     <- 1000;
C     <- 1;
sigma <- 0.4;
```

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supporting Information for other distributions which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 1$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered tightly around the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (red circle Fig 2b). This larger radius occurs because the addition of $Var(\gamma)$ increases the realised $\sigma$ of $M$. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's[@May1972; @Allesina2012] threshold of $\sigma\sqrt{SC} > 1$.

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from $2-32$ (see Supporting Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. Under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$.

<!--- Remember in Fig 3 that the line is showing the A1s that are stable *due to* var(gamma) -- sometimes A0s were stable, but they were destabilised by var(gamma) --->

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[fhalf]), digits = 3)` centred at $\gamma = -1.95$ and one small circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[shalf]), digits = 3)` centred at $\gamma = -0.05$. In a, the maximum real eigenvalaue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` were stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 1$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but clusters eigenvalues toward the distribution's centre (-1, 0). Black and red elipses in both panels show the circle centred on the distribution in panels a and b, respectively, which have a radius of $\sigma \sqrt{SC}$. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.
